---
title: "Data Visualization I: Base Graphics and ggplot2"
author: "Patrick Chester"
date: '`r format(Sys.time(), "%B %d %Y")`'
format: pdf
---

Last edited: 2/16/2022


# I. Base Graphics -

```{r}
library(MASS)
data(UScereal)
head(UScereal)
```



## A. Choosing the Right Graph for your Data

### One Continuous Variable
Key questions:
- What is the Distribution of Variable X?
- Is my Variable X normally distributed? Bimodal? Skewed?
- Are there any outliers in my variable?

Histogram

```{r}
hist(UScereal$calories, breaks = 15)
```

Boxplot

```{r}
boxplot(UScereal$calories, horizontal = TRUE)
```



### One Categorical (Discrete) Variable

Key question(s):
- Is the distribution across categories of Variable X evenly distributed?

```{r}
barplot(table(UScereal$shelf))
```




### Two Continuous Variables

Key questions:
- Is there a relationship between Variable X and Variable Y?
- If there is a relationship, is it linear? Quadratic? 

Scatterplot
```{r}
plot(x = UScereal$sugars, y = UScereal$calories)
```

Using a formula to create a scatterplot
```{r}
plot(calories ~ sugars, data = UScereal) # formula notation
```

Scatterplot matrix
```{r}
plot(UScereal[, c(2:8, 10)]) # scatterplot matrix
```




### One Continuous Variable and One Categorical Variable

Key question(s):
- Is the distribution of Variable Y, different across categories of Variable X?

```{r}
boxplot(sugars ~ shelf, data = UScereal)
```




### Two Continuous Variables and One Categorical Variable

Key questions:
- Is the relationship between Variable X and Y different across categories of Variable Z?

```{r}
plot(calories ~ sugars, data = UScereal, col = shelf)
```



## B. Graphing Elements

### Titles

*Tip: We want titles to meaningful, non-repetitive and include units when applicable*

Use the ylab, xlab and main arguments within the plot command

```{r}
plot(calories ~ sugars, data = UScereal, ylab = 'Calories',
     xlab = 'Sugars (grams)', main = 'Nutrition of a Single Cup of Cereal')
```


Use the title function after the plot has been generated
If using the title function, the ann argument inside plot must be set to FALSE

```{r}
plot(calories ~ sugars, data = UScereal, ann = FALSE)
title(main = 'Nutrition of a Single Cup of Cereal', ylab = 'Calories',
      xlab = 'Sugars (grams)') # add afterwards
```




### Legend

Use the legend function after the plot has been generated

```{r}
plot(calories ~ sugars, data = UScereal, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('green', 'red', 'black'), title = "Shelf")
```



### Point Shape and Color 

Tip: Changing color or shape of points can be used to represent the same dimension

The pch argument changes the shape of the points

```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
```



Set a color to a factor variable, and R will use default colors
```{r}
plot(calories ~ sugars, data = UScereal, pch = 19, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('green', 'red', 'black'), title = "Shelf")
```


Use a palette of defined colors
```{r}
palette(c('#e5f5f9', '#99d8c9', '#2ca25f'))
plot(calories ~ sugars, data = UScereal, pch = 19, col = shelf)
legend('topright', inset = .05, legend = c(3,2,1),
       fill = c('#e5f5f9', '#99d8c9', '#2ca25f'), title = "Shelf")
```


Label points with the text function

```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
text(UScereal$sugars, UScereal$calories, row.names(UScereal),
     col = "red", pos = 1, cex = .5)
```


The pos argument in the text funciton changes the position of the labels
```{r}
plot(calories ~ sugars, data = UScereal, pch = 15)
text(UScereal$sugars, UScereal$calories, UScereal$mfr, 
     col = "blue", pos = 2)
```


### Identify Outliers

1. Create a subset of datas which are the outliers, and then use that subset to label

```{r}
plot(calories ~ sugars, data = UScereal, pch = 19)
outliers <- UScereal[which(UScereal$calories > 300),]
text(outliers$sugars, outliers$calories, labels = row.names(outliers),pos = 1)
```

2. Or remove them by subsetting UScereal
```{r}
plot(calories ~ sugars, data = UScereal[UScereal$calories < 300, ], pch = 19)
```

### Size (text size, point size, label size etc..)

Use the cex argument family to change the size of any attribute
```{r}
plot(calories ~ sugars, data = UScereal, pch = 19, ann = FALSE, cex = 1.5)
outliers <- UScereal[which(UScereal$calories > 300),]
text(outliers$sugars, outliers$calories, pos = 1,
     labels = row.names(outliers), cex = .75)
title(main = 'Nutrition of a Single Cup of Cereal', ylab = 'Calories',
      xlab = 'Sugars (grams)', cex.main = 1.5, cex.lab = 1.0)
```


Combine Graphs into the Same Window
```{r}
par(mfrow = c(2, 2))

boxplot(calories ~ shelf, data = UScereal)
hist(UScereal$calories, breaks = 15)
boxplot(sugars ~ shelf, data = UScereal)
hist(UScereal$sugars, breaks = 15)

par(mfrow = c(1, 1)) # reset the matrix
```


### Exercise 1a. 
How can we improve this graph? From what we have learned above, implement at least 3 improvements to this graph.
A little more information about this dataset: The heart and body weights of samples of male and female cats used for digitalis experiments. The cats were all adult, over 2 kg body weight.

```{r}
data(cats)
head(cats)
plot(Hwt~Bwt, data = cats)
```
### Exercise 1b.
How can we improve this graph? From what we have learned above, implement at least 3 improvements to this graph. A little more information about this dataset: The heart and body weights of samples of male and Fisher's famours dataset measures the sepal and petal length and width for 3 species of Iris
```{r}
data(iris)
head(iris)
boxplot(Petal.Width ~ Species, data = iris)
```


# II. Advanced Graphics using ggplot2

```{r}
#install.packages('ggplot2')
library(ggplot2)
```

## A. The ggplot function

Base q plot function
```{r}
qplot(x = sugars, y = calories, color = as.factor(shelf), data = UScereal) 
```

Use the ggplot function to get started
```{r}
p1 <- ggplot(UScereal, aes(x = calories))
```


## B. Layers 

Add layers to the original ggplot object with a '+'
```{r}
p1 + geom_dotplot()

p1 + geom_density()

p1 + geom_histogram(binwidth = 10)
```

And we add more layers
```{r}
p1 + geom_histogram(binwidth = 10) + 
  xlab('Calories in a Single Cup') +
  ylab('Frequency') + 
  ggtitle('Distribution of Calories in US Cereals') + 
  theme_bw()
```

You can also add multiple geom_function layers to the same graph
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories, color = mfr))
```

Overlaying points colored by manufacturer
```{r}
p2  + geom_point() 
```

Overlaying fitted OLS regression lines by manufacturer
```{r}
p2  + 
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) 
```

To color by manufacturer - we can put the color in the ggplot function:
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories, color = mfr))

p2 + geom_point() 
```

Or inside the geom_point function:
```{r}
p2 <- ggplot(UScereal, aes(x = sugars, y = calories))

p2 + geom_point(aes(color = mfr)) 
```

## C. Aesthetics: x position, y position, size of elements, shape of elements, color of elements 
elements: geometric shapes such as points, lines, line segments, bars and text
geometries have their own aesthetics i.e. points have their own shape and size

We can scale colors manually using the appropriately named scale_color_manual function
```{r}
my_colors <- c('#9ebcda', '#8c96c6', '#8c6bb1', '#88419d', '#810f7c', '#4d004b')

p2 + geom_point(aes(color = mfr)) + 
  scale_color_manual(values = my_colors) 
```

### Adding Labels to points

Use the geom_text() layer
```{r}
p2 + geom_point(aes(color = mfr)) + 
  geom_text(aes(label = row.names(UScereal)), hjust = 1.1)
```

Changing point size
```{r}
p2 + geom_point(aes(color = mfr), size = 2) 
```

### Editing the legend

Use the scale_color_manual() layer, and the color argument in the labs() layer 
```{r}
p2 + geom_point(aes(color = mfr), size = 2) + 
  labs(color = 'Manufacturer') + 
  scale_color_manual(values = c('blue', 'green', 'purple', 'navyblue', 'red', 'orange'),  
                     labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina'))
```

### Changing text size

We can also change the size of legend and axis labels
```{r}
p2 + geom_point(aes(color = mfr), size = 2) + 
  labs(color = 'Manufacturer') + 
  scale_color_manual(values = c('blue', 'green', 'purple', 'navyblue', 'red', 'orange'),  
                     labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina')) + 
  theme(text = element_text(size = 15)) 
```

## D. Faceting  - divide a plot into subplots based on the values of one or more discrete variables

Tip: Use facets to help tell your story
Key questions:
- How is the distribution of sugar across different shelves?
- Are cereals with higher sugar content on lower shelves/at a child's eye level?

```{r}
p3 <- ggplot(UScereal, aes(x = sugars))

p3 + geom_histogram(binwidth = 4)
```

Each graph is in a separate row of the window
```{r}
p3 + geom_histogram(binwidth = 4) + 
  facet_grid(shelf ~ .)
```

Each graph is in a separate column of the window
```{r}
p3 + geom_histogram(binwidth = 4) + 
  facet_grid(. ~ shelf)
```

Finished product 
```{r}
p3 + geom_histogram(fill = '#3182bd', color = '#08519c', binwidth = 4) +
  facet_grid(shelf ~ .) + 
  theme(text = element_text(size = 15)) + 
  labs(title = 'Are Sugary Cereals on Lower Shelves?',
       x = 'Sugars (grams)', y = 'Count')
```

### Box Plots
```{r}
p4 <- ggplot(UScereal, aes(mfr, calories))
```

```{r}
p4 + geom_boxplot()
```


```{r}
p4 + geom_boxplot(outlier.shape = 8, outlier.size = 3, fill = '#3182bd') + 
  coord_flip() + 
  labs(x = 'Manufacturer', y = 'Calories') + 
  theme_bw() + 
  scale_x_discrete(labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina'))
```

# Add median value to boxplot

```{r}
#install.packages('dplyr')
library(dplyr)
p4_meds <- UScereal %>% 
  group_by(mfr) %>% 
  summarise(med = round(median(calories), 1))
```

```{r}
p4 + geom_boxplot(outlier.shape = 8, outlier.size = 3, fill = '#8c96c6') + 
  labs(x = 'Manufacturer', y = 'Calories') + 
  theme_bw() + 
  scale_x_discrete(labels = c('General Mills', 'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats', 'Ralston Purina')) + 
  geom_text(data = p4_meds, aes(x = mfr, y = med, label = med), size = 4, vjust = 1.4)
```


### Exercise 2a. Use ggplot2 to improve the graph below:
```{r}
data(cats)
head(cats)
plot(Hwt~Bwt, data = cats)
```


### Exercise 2b. Use ggplot2 to improve the graph below:
```{r}
data(iris)
head(iris)
boxplot(Petal.Width ~ Species, data = iris)
```

# III. Other Graphics

## Scatterplot Matrix
```{r}
#install.packages('GGally')
library(GGally)
ggpairs(UScereal[, c(2, 8, 9, 11)],
        upper = list(continuous = 'smooth', combo = 'facetdensity', discrete = 'blank') ,
        lower = list(continuous = 'cor', combo = 'box'))
```

Maps
http://bcb.dfci.harvard.edu/~aedin/courses/R/CDC/maps.html
http://rstudio.github.io/leaflet/
maps, choroplethr, 


# IV. Exercise Solutions

Below are just examples of cleaned up graphics. There are many solutions on how to improve these graphs

## Exercise 1a 

```{r}
palette(c('#fa9fb5', '#9ebcda'))
plot(Hwt~Bwt, data = cats, ylab  = 'Heart Weight (grams)', xlab = 'Body Weight (kg)', main = 'Measurements of Cats', pch = 16, col = Sex)
```

## Exercise 1b
```{r}
boxplot(Petal.Width ~ Species, data = iris, xlab = 'Petal Width (centimeters)', 
        main = 'Distribution of Petal Length by Species', pch = 8, horizontal = TRUE, col = 'lightgray')
```



## Exercise 2a
```{r}
e2a <- ggplot(cats, aes(Bwt, Hwt, color = Sex))
e2a + geom_point() + 
  theme_bw() + 
  labs(title = 'Measurements of Cats', x = 'Body Weight (kg)', y = 'Heart Weight (grams)')
```

## Exercise 2b
```{r}
e2b <- ggplot(iris, aes(Species, Petal.Width))
e2b + geom_boxplot(fill = '#addd8e') + 
  theme_bw() + 
  coord_flip() + 
  ggtitle('Distribution of Petal Length by Species' )
```


